---
title: "Chapter1"
format: html
---

Import necessary libraries

```{r}
library(tidyverse)
library(broom)
library(pwr)
```


### A basic experiment

`ToothGrowth` is a built-in R dataset from a study that examined the effect of three different doses of Vitamin C on the length of the odontoplasts, the cells responsible for teeth growth in 60 guinea pigs, where tooth length was the measured outcome variable. Let's load `ToothGrowth` into our environment, do some exploratory data analysis, and perform a `t-test` to get an idea of the purpose of a basic experiment.

```{r}
# Load the ToothGrowth dataset
data("ToothGrowth")

# View the first 6 rows of ToothGrowth
head(ToothGrowth)
```

```{r}
# Find mean len, median len, and standard deviation len with summarise()
ToothGrowth |> summarise(mean(len), median(len), sd(len))
```

```{r}
# Perform a two-sided t-test
t.test(x = ToothGrowth$len, alternative = "two.sided", mu = 18)
```

Given the high p-value, we fail to reject the null hypothesis that the mean of `len` is equal to 18. That is, we do not have evidence that it is different from 18 micrometers.

### Randomization

Randomization of subjects in an experiment helps spread any variability that exists naturally between subjects evenly across groups. For `ToothGrowth`, an example of effective randomization would be randomly assigning male and female guinea pigs into different experimental groups, ideally cancelling out any existing differences that naturally exist between male and female guinea pigs.

In the experiment that yielded the `ToothGrowth` dataset, guinea pigs were randomized to receive Vitamin C either through orange juice or ascorbic acid, indicated in the dataset by the `supp` variable. It is natural to wonder if there is a difference in tooth length by suuplement type - a question that a `t-test` can also answer.

Starting with this exercise, you should use `t.test()` and other modelling functions with formula notation:

`t.test(outcome ~ explanatory_variable, data = dataset)`

This can be read: "test `outcome` by `explanatory_variable` in my `dataset`." The default test for `t.test()` is a two-sided t-test.

```{r}
# Perform a t-test
ToothGrowth_ttest <- t.test(len ~ supp, data = ToothGrowth)

# Tidy ToothGrowth_ttest with tidy() from the broom package
tidy(ToothGrowth_ttest)
```

~~~{=comment}
this comment will be ignored
~~~

This comes after a comment that is being ignored.

### Replication

Recall that replication means you need to conduct an experiment with an adequate number of subjects to achieve an acceptable statistical power. Sample size and power will be discussed in more detail in this chapter.

Let's examine the `ToothGrowth` dataset to make sure they followed the principle of replication. We will use `dplyr` to do some exploratory data analysis (EDA).

```{r}
# Count number of observations for each combination of supp and dose
ToothGrowth |> 
  count(supp, dose)
```

The researchers seem to have tested each combination of `supp` and `dose` on 10 subjects each, which is low, but was deemed adequate for this experiment.

### Blocking

Though this is not true, suppose the supplement type is actually a nuisance factor we would like to control for by blocking, and we are actually only interested in the effect of the dose of Vitamin C on guinea pig tooth growth.

If we block by supplement type, we create groups that are more similar, in that they will have the same supplement type, allowing us to examine only the effect of dose on tooth length.

We will use the `aov()` function to examine this. `aov()` creates a linear regression model by calling `lm()` and examining results with `anova()` all in one function call. To use `aov()` we will still need functional notation, as with the randomization exercise, but this time the formula should be `len ~ dose + supp` to indicate we have blocked by supplement type.

```{r}
# Create a boxplot with geom_boxplot()
ggplot(data = ToothGrowth, aes(x = as_factor(dose), y = len)) +
  geom_boxplot()

```

```{r}
# Create ToothGrowth_aov
ToothGrowth_aov <- aov(len ~ dose + supp, data = ToothGrowth)

# Examine ToothGrowth_aov with summary()
summary(ToothGrowth_aov)
```

```{r}
# Alternative method
fit <- lm(len ~ dose + supp, data = ToothGrowth)
anova(fit)
```


### One sided vs. Two-sided tests

Recall in the first exercise that we tested to see if the mean of the guinea pigs' teeth in `ToothGrowth` was not equal to 18 micrometers. That was an example of a two-sided t-test: we wanted to see if the mean of `len` is some other number on either side of 18.

We can also conduct a one-sided t-test, explicitly checking to see if the mean is less than or greater than 18. Whether to use a one- or two-sided test usually follows from your research question. Does an intervention cause longer tooth growth? One-sided, greater than. Does a drug cause the test group to lose more weight? One-sided, less than. Is there a difference in mean test scores between two groups of students? Two-sided test.

```{r}
# Less than
t.test(x = ToothGrowth$len, alternative = "less", mu = 18)
```

```{r}
# Greater than
t.test(x = ToothGrowth$len, alternative = "greater", mu = 18)
```

It turns out the mean of `len` is actually very close to 18, so neither of these tests tells us much about the mean of tooth length.

### Power & Sample Size Calculations

One key part of designing an experiment is knowing the required sample size you will need to be able to test your hypothesis.

The `pwr` package provides a handy function, `pwr.t.test()`, which will calculate that for you. However, you do need to know your desired significance level (often 0.05), if the test is one- or two-sided, if the data is from one sample, two samples, or paired, the effect size, and the power. Some of this information will be given to you or can be reasoned from the design.

A power or sample size calculation is usually different each time you conduct one, and the details of the calculation strongly depend on what kind of experiment you are designing and what your end goals are.

```{r}
# Calculate power
pwr.t.test(n = 100, 
           d = 0.35, 
           sig.level = 0.1, 
           type = "two.sample", 
           alternative = "two.sided", 
           power = NULL)
```

```{r}
# Calculate sample size
pwr.t.test(n = NULL,
           d = 0.25,
           sig.level = 0.05,
           type = "one.sample",
           alternative = "greater",
           power = 0.8)
```

Because this sample size calculation was for a one-sided test, we only need 101 subjects, not 101 in each group. As you design experiments in the future, the `pwr` package includes functions for calculating power and sample size for a variety of different tests, including ANOVA.


